Discrete Time Markov Model

### 2025-09-30 analysis/discrete-markov.qmd
Updating analysis scripts
Adding scripts, files, and manuscript :fire:

# Data -------------------------------------------------------------------------
data <- read.csv(here::here("analysis/data/data_apathy.csv"))
```

@@

```{r covariates}
cols <- c(
  "ID", "Gender", "Age", "Education_tri",
  paste0("apathy_", seq(2016, 2022, by = 2)),
  "Total_p")
```

@@

```{r extract-years}
data_stack <- data |> 
  extract_years(years = seq(2016, 2022, by = 2), impute = TRUE, absorbing = TRUE) |>
  na.omit()

@@

```{r long_data}
data_stack <- data_stack |>
  inner_join(data[, cols], by = "ID") |>
  # Removing baseline age below 60
  filter(Age - (2022 - 2016) >= 60) |>
  pivot_and_factorise(time_vary = TRUE) |>
  group_by(ID) |>
  # ^ Make age time varying
  mutate(Age = Age - (2022 - as.numeric(as.character(wave))))

@@

```{r stack-data}
data_stack <- data_stack |>
  mutate(status_prev = lag(status), .after = status) |>
  filter(!is.na(Total_p)) |>
  ungroup() |> 
  filter(wave != 2016)

@@

# Transition frequencies

We will now calculate the transition frequencies between cognitive states for each time period. We will use the `create_transition_table()` function to create a transition table for each time period. This will be done with the help of the `map()` function from the `purrr` package. Finally, to combine all the transition tables into one dataset, we will use the `create_transition_dataset()` function.

```{r transition-tables}
# Creating a table dataset
table_data <- data |>
  extract_years(seq(2016, 2022, by = 2)) |>
  rename_with(~ gsub("cogfunction", "HRS_", .)) |>
  mutate(
    across(c(HRS_2016:HRS_2022), ~ factor(.x, levels = c("Normal Cognition", "MCI", "Dementia")))
  )

# Creating transition frequencies ---------------------------------------------
## These are the time periods we are interested in
time_periods <- list(
  c("2016", "2018"),
  c("2018", "2020"), 
  c("2020", "2022")
)

# Applying function
transition_results <- purrr::map(time_periods, ~ create_transition_table(.x[1], .x[2]))
names(transition_results) <- purrr::map_chr(time_periods, ~ paste(.x[2], .x[1], sep = "-"))

### Creating one dataset
transition_frequencies <- time_periods |>
  create_transition_dataset(transition_results = transition_results)

head(transition_frequencies)
```

## Observed transition matrix

The probability distribution of transitions from one state to another can be represented into a transition matrix $P = (p_{ij})_{i,j}$ where each element of position $(i, j)$ represents the transition probability $p_{ij}$.

In order to create this matrix we will use both the `create_transitions()` and `observed_transition_matrix()` functions.

```{r observed-matrix}
## Creating observed transition matrix -----------------------------------------
transition_matrix_observed <- data |>
  extract_years(seq(2016, 2022, by = 2)) |>
  create_transitions() |>
  observed_transition_matrix()

transition_matrix_observed
```

## Visualisation

Let's visualize both the transition frequencies (@fig-frequencies) and matrix (@fig-matrix).

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| fig-height: 10
#| label: fig-frequencies
#| fig-cap: "Transition frequencies between cognitive states for each time period."

fig_1 <- transition_frequencies |> tran_stack_graph()
fig_2 <- transition_frequencies |> tran_heat_map()

fig_1 / fig_2
```

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| label: fig-matrix
#| fig-cap: "Observed transition matrix between cognitive states for the years 2016 - 2022."

fig_3 <- transition_matrix_observed |>
  reshape_matrix() |>
  plot_transition_matrix()

fig_3
```

# Modelling

## Markov Process Fundamentals

@@

### Transition Probability Matrix

For our three-state system (Normal Cognition, Mild Cognitive Impairment [MCI], Dementia), the transition matrix $P$ captures all possible transition probabilities:

$$
P = \begin{bmatrix} 
p_{11} & p_{12} & p_{13} \\
p_{21} & p_{22} & p_{23} \\
p_{31} & p_{32} & p_{33} \\
\end{bmatrix}
$$ {#eq-t-matrix}

where:

- $p_{ij} = P(X_{t+1} = j \vert X_t = i)$ represents the probability of transitioning from state $i$ to state $j$.
- Each row sums to 1, $\sum^3_{j=1} p_{ij} = 1 \quad  \forall_i \in \{1, 2, 3\}$

@@

$$
log \left( \frac{P(Y = j \vert x)}{P(Y = k \vert x)} \right) = \beta_{j0} + \beta_j^Tx \qquad \text{for } j = 1, \dots K-1
$$ {#eq-multinomial}

@@

### Time-Homogeneous Approach

Our implementation assumes time-homogeneous transitions, but the framework can be extended to time-varying probabilities:

$$
P^{(t)} = \begin{bmatrix} 
p_{11}^{(t)} & p_{12}^{(t)} & \cdots & p_{1n}^{(t)} \\
p_{21}^{(t)} & p_{22}^{(t)} & \cdots & p_{2n}^{(t)} \\
\vdots & \vdots & \ddots & \vdots \\
p_{n1}^{(t)} & \cdots & \cdots & p_{nn}^{(t)} \\
\end{bmatrix}
$$ {#eq-non-stationary}

@@

We model Dementia as an absorbing state:

@@

## Stationary model

We estimate three progressively complex stationary models using `nnet::multinom()`

```{r model-fitting}
#| code-fold: true
#| code-summary: "Check out my models"
#| 
# Baseline model (only gender; reference: NC)
fit_1a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression, 
  family = multinomial, 
  data = data_stack, trace = FALSE)

# Baseline model (only gender; reference: MCI)
fit_1b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression, 
  family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# With procrastination (reference: NC)
fit_2a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p, 
  family = multinomial, 
  data = data_stack, trace = FALSE)

# With procrastination (reference: MCI)
fit_2b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p, 
  family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# Full model with previous state (reference NC)
fit_3a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev, 
  family = multinomial, 
  data = data_stack, trace = FALSE)

# Full model with previous state (reference MCI)
fit_3b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev, 
  family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# Full model with interaction (reference NC)
fit_4a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev, 
  family = multinomial, 
  data = data_stack, trace = FALSE)

# Full model with interaction (reference MCI)
fit_4b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev, 
  family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)
```

### Model comparison

We evaluate model improvement (@tbl-stationary-comparison) using likelihood ratio tests:

$$
D = -2 \times \ell_{\text{reduced}} - \ell_{\text{full}}
$$ {#eq-liklihood}

```{r}
#| label: tbl-stationary-comparison
#| tbl-cap: "Likelihood ratio test for stationary models"
#| tbl-cap-location: bottom

anova(fit_1a, fit_2a, fit_3a, fit_4a) |>
  mutate(
    Model = c("Baseline", 
              "Procrastination", 
              "Procrastination + Previous Status",
              "Interaction Model"),
    `Resid. Dev`= round(`Resid. Dev`, digits = 3),
    `LR stat.` = round(`LR stat.`, digits = 3),
    `Pr(Chi)` = round(`Pr(Chi)`, digits = 3)
    ) |>
  DT::datatable(
    options = list(
      pageLength = 4,
      dom = "t",
      ordering = FALSE,
      columnDefs = list(list(className = "dt-center", targets = "_all"))
    ),
    rownames = FALSE
  )
```

::: {.panel-tabset .nav-pills}
## Non-interaction model

@@

Below we present the estimated parameters for the stationary model (fit 3) (@tbl-stationary-output). We will transform the estimates to odds ratios and colour them based on significance.

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| label: tbl-stationary-output
#| tbl-cap: "Estimated parameters for the stationary model"
#| tbl-cap-location: bottom

stationary_results <- rbind(tidy_output(fit_3a), tidy_output(fit_3b)) |>
  mutate(
    # Mutating to factors
    transition = factor(
      transition, 
      levels = c("NC - MCI", "MCI - NC", 
                 "NC - Dementia", "MCI - Dementia")),
    term = factor(
      term, 
      levels = c("Being female", "Age", "High school degree vs. No education",
                 "Further education vs. No education", "Depression",
                 "Procrastination (2020)", "Previous state: MCI", 
                 "Previous state: Dementia"))
    )

stationary_results |>
  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
  DT::datatable(
    options = list(
      pageLength = 6,
      dom = "tip",
      ordering = FALSE,
      columnDefs = list(list(className = "dt-center", targets = "_all"))
    ),
    rownames = FALSE
  )
```

Let's visualize the estimated odds ratios (@fig-odds-ratio) and the predicted transition probabilities (@fig-model-predictions) for the stationary model .

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| label: fig-odds-ratio
#| fig-cap: "Estimated odds ratios for the stationary model."

fig_4a <- stationary_results |>
  filter(!term %in% c("Previous state: MCI", "Previous state: Dementia")) |>
  ggplot(aes(x = estimate, y = transition, colour = colour)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  ggstance::geom_pointrangeh(
    aes(xmin = conf.low, xmax = conf.high),
    position = ggstance::position_dodgev(height = 0.5),
    size = 1.25,
    fatten = 3) +
  scale_colour_manual(values = c(
    "Positive" = "#0072B2", 
    "Negative" = "#E69F00", 
    "NS"       = "#B2BEB5")) +
  labs(title = "Odds ratios (stationary model)",
       x = "Odds Ratio", y = "Predictor") +
  guides(colour = "none") +
  facet_wrap(~ term, scales = "free_x") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 10, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank())

fig_4a
```

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| label: fig-model-predictions
#| fig-cap: "Predicted transition probabilities from the stationary model."
#| 
## Making a prediction dataset -------------------------------------------------
pred_data <- expand.grid(
  Gender = factor(0),
  Age = mean(data_stack$Age),
  Education_tri = factor(0),
  Depression = mean(data_stack$Depression, na.rm = TRUE),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200))

## Plotting predictions
fig_4b <- pred_data |>
  modelr::add_predictions(model = fit_3a, var = "pred", type = "probs") |>
  tidy_predictions() |>
  plot_predictions_stationary(variable = Total_p, x_axis = "Total Procrastination")

fig_4b
```

### Estimated transition matrix

Finally, we can compare the observed and predicted transition matrices. We will use the estimated probabilities from the model to fill in the transition matrix.

```{r estimated-matrix}
# Get all unique states
names <- c("Normal Cognition", "MCI", "Dementia")
states <- sort(unique(data_stack$status))
n_states <- length(states)

# Create empty transition matrix
transition_matrix_estimated <- matrix(
  0, nrow = n_states, ncol = n_states,
  dimnames = list(paste("From", names), paste("To", names))
  )

# Getting estimated probabilities
estimated_probs <- expand.grid(
  Gender = factor(0),
  Age = mean(data_stack$Age),
  Education_tri = factor(0),
  Depression = mean(data_stack$Depression, na.rm = TRUE),
  Total_p = mean(data_stack$Total_p),
  status_prev = states) |>
  modelr::add_predictions(model = fit_3a, var = "pred", type = "probs")

# Filling in matrix
for(i in 1:n_states) {
  transition_matrix_estimated[i, ] <- estimated_probs$pred[i, ]
}

transition_matrix_estimated |> round(digits = 3)
```

Let's plot this and then compare with our observed matrix

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| label: fig-matrix-comparison
#| fig-cap: "Comparison of observed and estimated transition matrices."

fig_5 <- transition_matrix_estimated |>
  reshape_matrix() |>
  plot_transition_matrix(observed = FALSE)

fig_3 + fig_5 + plot_layout(axis_titles = "collect", guides = "collect")
```

## Interaction model

Below we present the estimated parameters for the interaction model (fit 4; @tbl-interaction-output). We will transform the estimates to odds ratios and colour them based on significance.

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| label: tbl-interaction-output
#| tbl-cap: "Estimated parameters for the interaction model"
#| tbl-cap-location: bottom

interaction_results <- rbind(tidy_output(fit_4a), tidy_output(fit_4b)) |>
  mutate(
    # Mutating to factors
    transition = factor(
      transition, 
      levels = c("NC - MCI", "MCI - NC", 
                 "NC - Dementia", "MCI - Dementia")),
    term = factor(
      term,
      levels = c("Being female", "Age", 
                 "High school degree vs. No education",
                 "Further education vs. No education", "Depression",
                 "Procrastination (2020)",  
                 "Age x Procrastination", "Previous state: MCI", 
                 "Previous state: Dementia")),
    )

@@

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| fig-height: 10
#| label: fig-odds-ratio-interaction
#| fig-cap: "Estimated odds ratios for the interaction model."

@@

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| warning: false
#| fig-width: 12
#| fig-height: 10
#| label: fig-interaction-predictions
#| fig-cap: "Predicted transition probabilities from the interaction model."

@@

# Plotting interaction effect ----------------------------------------------
pred_data |>
  # Adding predictions
  modelr::add_predictions(model = fit_4a, var = "pred", type = "probs") |>
  tidy_predictions() |>
  filter(status_prev %in% c("Normal Cognition", "MCI") & 
         status %in%  c("Normal Cognition", "MCI")) |>
  ggplot(aes(x = Total_p, y = as.numeric(Age))) +
  # Adding heat mat with contour lines
  geom_raster(aes(fill = prob), alpha = 0.7) +
  geom_contour(aes(z = prob), colour = "black", size = 0.3) +
  # Adding orginal data points
  geom_point(
    data = data_plot, aes(x = Total_p, y = Age), size = 1.75, alpha = 0.3) +
  # Customising plot
  scale_x_continuous(
    breaks = seq(0, 60, by = 10),
    sec.axis = sec_axis(~ ., name = "Previous State",
                        breaks = NULL, labels = NULL)) +
  scale_y_continuous(
    breaks = seq(50, 100, by = 5),
    sec.axis = sec_axis(~ ., name = "Current State",
                        breaks = NULL, labels = NULL)) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    title = "Predicted transition probabilities from the interaction model",
    x = "Total Procrastination", y = "Age", 
    fill = expression(hat(p))) +
  facet_grid(status ~ status_prev) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 10, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.title = element_text(size = 10, face = "bold"))

@@

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| warning: false
#| fig-width: 12
#| fig-height: 8
#| label: fig-interaction-age-predictions
#| fig-cap: "Predicted transition probabilities for different age cohorts."

# Create predictions with age info
age_cohots <- expand.grid(
  Gender = factor(0),
  Age = c(62:97),
  Education_tri = factor(0),
  Depression = mean(data_stack$Depression, na.rm = TRUE),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200)) |>
  modelr::add_predictions(model = fit_4a, var = "pred", type = "probs") |>
  tidy_predictions() |>
  filter(status_prev %in% c("Normal Cognition", "MCI") &
         status %in%  c("Normal Cognition", "MCI")) |>
  mutate(label = case_when(
    Age %in% c(70, 80, 90) ~ as.character(Age), 
    TRUE ~ "Other"
  ))

@@

# Plotting ---------------------------------------------------------------------
ggplot(data = age_cohots, aes(x = Total_p, y = prob, 
                              group = interaction(Age, status_prev, status),
                              colour = label)) +
  geom_line(aes(linewidth = label)) + 
  ggrepel::geom_text_repel(
    data = label_data,
    aes(label = paste0("Age ", Age), colour = as.character(Age)),
    direction = "y", nudge_x = 4, hjust = 0,
    segment.size = 0.5, segment.colour = "grey50",
    size = 3, show.legend = FALSE) +
  scale_x_continuous(
    breaks = seq(0, 60, by = 10),
    sec.axis = sec_axis( ~., name = "Previous State", breaks = NULL, labels = NULL)) +
  scale_y_continuous(sec.axis = sec_axis( ~ ., name = "Current State", breaks = NULL, labels = NULL)) +
  scale_colour_manual(values = c(
    "70"    = "#E69F00", 
    "80"    = "#56B4E9", 
    "90"    = "#009E73",
    "Other" = "grey80")) +
  scale_linewidth_manual(values = c(
    "70"    = 1.5, 
    "80"    = 1.5, 
    "90"    = 1.5,
    "Other" = 0.5)) +
  labs(
    title = "Predicted transition probabilities for different age cohorts",
    x = "Procrastination", y = "Probability", colour = NULL) +
  facet_grid(status ~ status_prev, labeller = labeller(
    status = c("Normal Cognition" = "Normal Cognitive Function"),
    status_prev = c("Normal Cognition" = "Normal Cognitive Function")
  )) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 10, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.title = element_text(size = 10, face = "bold"),
    legend.position = "none")
```

:::

## Non-stationary model

We can also estimate a non-stationary model (@eq-non-stationary) by incorporating time `(wave)` into our model. This model allows for time-varying transition probabilities $p^{(t)}_{ij}$.

We estimate three progressively complex non-stationary models:

- Additive time effects `(fit_4)`: Baseline covariates + wave
- State-specific time effects `(fit_5)`: Interaction between previous state and wave
- Full time interactions `(fit_6)`: All covariates interacting with wave

```{r non-stationary}
#| code-fold: true
#| code-summary: "Check out my models"

## Small processing to fix wave column
data_stack <- data_stack |>
  mutate(wave = case_when(
    wave == "2016" ~ 1,
    wave == "2018" ~ 2,
    wave == "2020" ~ 3,
    wave == "2022" ~ 4
  ))

# Model 5: Additive time effects
fit_5a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) +  status_prev + wave, family = multinomial, 
  data = data_stack, trace = FALSE)

fit_5b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev + wave, family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# Model 6: State-specific time effects
fit_6a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + (status_prev * wave), family = multinomial, 
  data = data_stack, trace = FALSE)

fit_6b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + (status_prev * wave), family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# Model 7: Full time interactions
fit_7a <- nnet::multinom(
  status ~ (Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev) * wave, family = multinomial, 
  data = data_stack, trace = FALSE)

fit_7b <- nnet::multinom(
  status ~ (Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev) * wave, family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)
```

### Model comparison

Again, we will evaluate model improvement using likelihood ratio tests (@eq-liklihood; @tbl-non-stationary-comparison). However, this time, we will also compare the best fitted non-stationary models `fit_3`.

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| label: tbl-non-stationary-comparison
#| tbl-cap: "Likelihood ratio test for non-stationary models"
#| tbl-cap-location: bottom

# Fit statistics
models <- list(
  "Interaction" = fit_4a,
  "Additive Time" = fit_5a,
  "State-Time Interaction" = fit_6a,
  "Full Interactions" = fit_7a
)

# Create comparison table
comparison_table <- purrr::map_dfr(models, broom::glance, .id = "Model") |>
  select(edf:AIC) |>
  mutate(across(where(is.numeric), \(x) round(x, digits = 1)))

# Outputting as table
anova(fit_4a, fit_5a, fit_6a, fit_7a) |>
  mutate(
    Model = c("Interaction", "Additive", "State-Time Interactions", "Full Interactions"),
    `Resid. Dev` = round(`Resid. Dev`, digits = 3),
    `LR stat.` = round(`LR stat.`, digits = 3),
    `Pr(Chi)` = round(`Pr(Chi)`, digits = 3)
    ) |>
  cbind(comparison_table) |>
  DT::datatable(
    options = list(
      pagelength = 4,
      dom = "t",
      ordering = FALSE,
      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    ),
    rownames = FALSE
  )
```

**This is interesting!! Looks like adding in time as an additive term improved things. However, having full time interactions is better than having state-time interactions**

::: {.panel-tabset .nav-pills}
## Additive Model

Key coefficients from the selected non-stationary model `fit_4`

```{r non-stationary-output}
additive_results <- rbind(tidy_output(fit_5a), tidy_output(fit_5b))

additive_results |>
  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
  DT::datatable(
    options = list(
      pageLength = 6,
      ordering = FALSE,
      columnDefs = list(list(className = "dt-center", targets = "_all"))
    ),
    rownames = FALSE,
    colnames = c("Transition", "Predictor", "Estimate", "Standard Error",
                 "Statistic", "p-val", "Lower CI", "Upper CI")
  )
```

Let's visualize the predicted transition probabilities from the non-stationary model.

- @fig-non-stationary-age shows the predicted transition probabilities for age
- @fig-non-stationary-depression shows the predicted transition probabilities for depression
- @fig-non-stationary-education shows the predicted transition probabilities for per education level
- @fig-non-stationary-procrastination shows the predicted transition probabilities for procrastination

::: {.panel-tabset .nav-pills}

## Age

```{r}
#| code-fold: true
#| code-summary: "Check our my code"
#| fig-width: 10
#| fig-height: 8
#| label: fig-non-stationary-age
#| fig-cap: "Predicted transition probabilities from the non-stationary model (for age)."
#|

create_additive_predictions(data = data_stack, model = fit_5a, var = "Age", var_seq = seq(50, 97, by = 5)) |>
  plot_additive_predictions(var = "Age", x_label = "Age", subtitle = "By Age")

## Depression

```{r}
#| code-fold: true
#| code-summary: "Check our my code"
#| fig-width: 10
#| fig-height: 8
#| label: fig-non-stationary-depression
#| fig-cap: "Predicted transition probabilities from the non-stationary model (for depression)."
#|

create_additive_predictions(data = data_stack, model = fit_5a, var = "Depression", var_seq = seq(0, 8, by = 1)) |>
  plot_additive_predictions(var = "Depression", x_label = "Depression", subtitle = "By depression")

## Education Level

```{r}
#| code-fold: true
#| code-summary: "Check our my code"
#| fig-width: 10
#| fig-height: 8
#| label: fig-non-stationary-education
#| fig-cap: "Predicted transition probabilities from the non-stationary model (for having a high school degree)."

create_additive_predictions(data = data_stack, model = fit_5a, var = "Education_tri", var_seq = factor(c(0:2)), hold_constant = list(Gender = factor(0))) |>
  plot_additive_predictions(var = "Education_tri", x_label = "Education", subtitle = "By education level") +
  geom_point(size = 2) +
  scale_x_continuous(breaks = c(0, 1, 2), labels = c("No Education", "High School", "Further Education")) +
  theme(axis.text.x = element_text(size = 8, angle = 60, vjust = 0.90, hjust = 1))
```

## Procrastination

```{r}
#| code-fold: true
#| code-summary: "Check our my code"
#| fig-width: 10
#| fig-height: 8
#| label: fig-non-stationary-procrastination
#| fig-cap: "Predicted transition probabilities from the non-stationary model (for procrastination)."
#|

create_additive_predictions(data = data_stack, model = fit_5a, var = "Total_p", var_seq = seq(0, 60, by = 1)) |>
  plot_additive_predictions(var = "Total_p", x_label = "Procrastination", subtitle = "By procrastination")

## Interaction

```{r}
#| code-fold: true
#| code-summary: "Check our my code"
#| fig-width: 12
#| fig-height: 10
#| label: fig-non-stationary-interaction
#| fig-cap: "Predicted transition probabilities from the non-stationary model (interaction effect)."

# Making prediction dataset ----------------------------------------------------
pred_data <- expand.grid(
  Gender = factor(0),
  Age = c(62:97),
  Education_tri = factor(0),
  Depression = mean(data_stack$Depression),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200),
  wave = c(2:4)
) |>
  modelr::add_predictions(model = fit_5a, var = "pred", type = "prob") |>
  tidy_predictions() |>
  filter(status_prev != "Dementia", status != "Dementia") |>
  mutate(label = case_when(
    Age %in% c(70, 80, 90) ~ as.character(Age), 
    TRUE ~ "Other"
  ))

# For adding age labels --------------------------------------------------------
label_data <- pred_data |>
  filter(Age %in% c(70, 80, 90)) |>
  group_by(status_prev, Age) |>
  filter(Total_p == max(Total_p)) |>
  ungroup()

>>>>

# Data -------------------------------------------------------------------------
data <- read.csv(here::here("analysis/data/data.csv"))
```

@@

```{r}
#| label: covariates
#| 
cols <- c(
  "ID", "Gender", "Age", "Education_tri",
  paste0("apathy_", seq(2016, 2022, by = 2)),
  "Total_p")
```

@@

```{r}
#| label: extract-years
#| collapse: true
data_stack <- data |> 
  extract_years(years = seq(2016, 2022, by = 2), impute = TRUE, absorbing = TRUE) |>
  na.omit()

@@

```{r}
#| label: long_data
#| collapse: true
data_stack <- data_stack |>
  # Join covariate data
  inner_join(data[, cols], by = "ID") |>
  # Removing baseline age below 60
  filter(Age - (2022 - 2016) >= 60) |>
  pivot_and_factorise(time_vary = TRUE) |>
  group_by(ID) |>
  # Make age time varying
  mutate(Age = Age - (2022 - as.numeric(as.character(wave))))

@@

```{r}
#| label: stack-data
#| collapse: true
#| 
data_stack <- data_stack |>
  mutate(status_prev = lag(status), .after = status) |>
  filter(!is.na(Total_p)) |>
  ungroup() |> 
  filter(wave != 2016)

@@

# Modelling
## Markov Process Fundamentals

@@

@@

$$
log \left( \frac{P(Y = j \vert x)}{P(Y = k \vert x)} \right) = \beta_0 + \beta_j^Tx \qquad \text{for } j = 1, \dots K-1
$$ {#eq-multinomial}

@@

@@

However, we model Dementia as an absorbing state:

@@

## Our model

As mentioned, we estimate our Markov model using `nnet::multinom()`

```{r}
#| label: model-fit

fit_a <- nnet::multinom(
  status ~ Gender + Education_tri + Depression + (Total_p * Age) + status_prev, 
  family = multinomial, 
  data = data_stack, trace = FALSE)

fit_b <- nnet::multinom(
  status ~ Gender + Education_tri + Depression + (Total_p * Age) + status_prev, 
  family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)
```

@@

Below we present the estimated parameters for the interaction model (@tbl-interaction-output). We will transform the estimates to odds ratios and colour them based on significance.

```{r}
#| label: tbl-interaction-output
#| tbl-cap: "Estimated parameters for the interaction model"
#| tbl-cap-location: bottom
#| code-fold: true
#| code-summary: "Check out my code"

interaction_results <- rbind(tidy_output(fit_a), tidy_output(fit_b)) |>
  mutate(
    # Mutating to factors
    transition = factor(
      transition, 
      levels = c("NC - MCI", "MCI - NC", 
                 "NC - Dementia", "MCI - Dementia")),
    term = factor(
      term,
      levels = c("Being female", "Age", 
                 "High school degree vs. No education",
                 "Further education vs. No education", "Depression",
                 "Procrastination (2020)",  
                 "Age x Procrastination", "Previous state: MCI", 
                 "Previous state: Dementia")),
    )

@@

```{r}
#| label: fig-odds-ratio-interaction
#| fig-width: 12
#| fig-height: 10
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-cap: "Estimated odds ratios for the interaction model."
#| message: false
#| warning: false

@@

```{r}
#| label: fig-interaction-predictions
#| fig-width: 12
#| fig-height: 10
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-cap: "Predicted transition probabilities from the interaction model."
#| warning: false

@@

# Plotting interaction effect ----------------------------------------------
pred_data |>
  # Adding predictions
  modelr::add_predictions(model = fit_a, var = "pred", type = "probs") |>
  tidy_predictions() |>
  filter(status_prev %in% c("Normal Cognition", "MCI") & 
         status %in%  c("Normal Cognition", "MCI")) |>
  ggplot(aes(x = Total_p, y = as.numeric(Age))) +
  # Adding heat mat with contour lines
  geom_raster(aes(fill = prob), alpha = 0.7) +
  geom_contour(aes(z = prob), colour = "black", size = 0.3) +
  # Adding orginal data points
  geom_point(
    data = data_plot, aes(x = Total_p, y = Age), size = 1.75, alpha = 0.3) +
  # Customising plot
  scale_x_continuous(
    breaks = seq(0, 60, by = 10),
    sec.axis = sec_axis(~ ., name = "Previous State",
                        breaks = NULL, labels = NULL)) +
  scale_y_continuous(
    breaks = seq(50, 100, by = 5),
    sec.axis = sec_axis(~ ., name = "Current State",
                        breaks = NULL, labels = NULL)) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    title = "Predicted transition probabilities from the interaction model",
    x = "Total Procrastination", y = "Age", 
    fill = expression(hat(p))) +
  facet_grid(status ~ status_prev) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 10, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.title = element_text(size = 10, face = "bold"))

@@

```{r}
#| label: fig-interaction-age-predictions
#| fig-width: 12
#| fig-height: 8
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-cap: "Predicted transition probabilities for different age cohorts."
#| warning: false

# Create predictions with age info
age_cohots <- expand.grid(
  Gender = factor(0),
  Age = c(62:97),
  Education_tri = factor(0),
  Depression = mean(data_stack$Depression, na.rm = TRUE),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200)) |>
  modelr::add_predictions(model = fit_a, var = "pred", type = "probs") |>
  tidy_predictions() |>
  filter(status_prev %in% c("Normal Cognition", "MCI") &
         status %in%  c("Normal Cognition", "MCI")) |>
  mutate(label = case_when(
    Age %in% c(70, 80, 90) ~ as.character(Age), 
    TRUE ~ "Other"
  ))

@@

# Plotting ---------------------------------------------------------------------
ggplot(data = age_cohots, aes(x = Total_p, y = prob, 
                              group = interaction(Age, status_prev, status),
                              colour = label)) +
  geom_line(aes(linewidth = label)) + 
  ggrepel::geom_text_repel(
    data = label_data,
    aes(label = paste0("Age ", Age), colour = as.character(Age)),
    direction = "y", nudge_x = 4, hjust = 0,
    segment.size = 0.5, segment.colour = "grey50",
    size = 3, show.legend = FALSE) +
  scale_x_continuous(
    breaks = seq(0, 60, by = 10),
    sec.axis = sec_axis( ~., name = "Previous State", breaks = NULL, labels = NULL)) +
  scale_y_continuous(sec.axis = sec_axis( ~ ., name = "Current State", breaks = NULL, labels = NULL)) +
  scale_colour_manual(values = c(
    "70"    = "#E69F00", 
    "80"    = "#56B4E9", 
    "90"    = "#009E73",
    "Other" = "grey80")) +
  scale_linewidth_manual(values = c(
    "70"    = 1.5, 
    "80"    = 1.5, 
    "90"    = 1.5,
    "Other" = 0.5)) +
  labs(
    # title = "Predicted transition probabilities for different age cohorts",
    x = "Procrastination", y = "Probability", colour = NULL) +
  facet_grid(status ~ status_prev, labeller = labeller(
    status = c("Normal Cognition" = "Normative Cognitive Function"),
    status_prev = c("Normal Cognition" = "Normative Cognitive Function")
  )) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 10, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.title = element_text(size = 10, face = "bold"),
    legend.position = "none")
```
